DCMIP2016, Part 1: Models and Equation Sets

### 2016-11-11 DCMIP2016-Part1-Models.tex
- Added about half of the equations section

\begin{table}[h]
\caption{A list of physical constants used in this document.} \label{tab:PhysicalConstants}
%\ \\
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lll}
\hline Constant & Description & Value \\
\hline $a_{\tiny \mbox{ref}}$ & Radius of the Earth & $6.37122 \times 10^{6}\ \mbox{m}$ \\
$\Omega_{\tiny \mbox{ref}}$ & Rotational speed of the Earth & $7.292\ \times 10^{-5}\ \mbox{s}^{-1}$ \\
%$a$ & Scaled radius of the Earth & $a_{\tiny \mbox{ref}} / X$ \\
%$\Omega$ & Scaled rotational speed of the Earth & $\Omega_{\tiny \mbox{ref}} \cdot X$ \\
$g_c$ & Gravitational acceleration & $9.80616\ \mbox{m}\ \mbox{s}^{-2}$ \\
$p_0$ & Reference pressure & $1000\ \mbox{hPa}$ \\
$c_p$ & Specific heat capacity of dry air at constant pressure & $1004.5\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
$c_v$ & Specific heat capacity of dry air at constant volume & $717.5\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
$R_d$ & Gas constant for dry air & $287.0\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
$R_\nu$ & Gas constant for water vapor & $461.5$ J kg$^{-1}$ K$^{-1}$ \\
$\kappa$ & Ratio of $R_d$ to $c_p$ & $2/7$ \\
$\varepsilon$ & Ratio of $R_d$ to $R_\nu$ & $0.622$ \\
$M_v$ & Constant for virtual temperature conversion & $0.608$ \\
$\rho_{water}$ & Reference density of water & 1000 kg m$^{-3}$ \\
\hline 
\end{tabular*}

@@

\section{Equation Sets} \label{sec:EquationSets}

{\color{red}[ALL] Include the continuous equation set that you use for your model here.}

@@

\noindent with horizontal velocity ${\bf V}_h$, 
generalized vertical velocity $\dot{ \zeta }=\frac{d\zeta}{dt}$, 
vertical velocity $w$,
geopotential $\phi$, 
virtual temperature $T_v$, 
pressure $p$ and
hydrostatic pressure $\pi$. The variable 
$\mu$ is a measure of departure from hydrostatic balance, 
$s$ is related to the surface pressure $\pi_s$ and $B$ is a metric term. The last two variables are defined in (\ref{eq:gem_zeta}).
The equations contain a total time derivative $\frac{ {d } }{dt}$ that expresses the change in time while following the air parcel.

@@

\subsection{Tracer transport}

@@

>>>>

\begin{table}[h]
\caption{A list of physical constants used in this document.} \label{tab:PhysicalConstants}
%\ \\
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lll}
\hline Constant & Description & Value \\
\hline $a_{\tiny \mbox{ref}}$ & Radius of the Earth & $6.37122 \times 10^{6}\ \mbox{m}$ \\
$\Omega_{\tiny \mbox{ref}}$ & Rotational speed of the Earth & $7.292\ \times 10^{-5}\ \mbox{s}^{-1}$ \\
%$a$ & Scaled radius of the Earth & $a_{\tiny \mbox{ref}} / X$ \\
%$\Omega$ & Scaled rotational speed of the Earth & $\Omega_{\tiny \mbox{ref}} \cdot X$ \\
$g_c$ & Gravitational acceleration & $9.80616\ \mbox{m}\ \mbox{s}^{-2}$ \\
$p_0$ & Reference pressure & $1000\ \mbox{hPa}$ \\
$c_{pd}$ & Specific heat capacity of dry air at constant pressure & $1004.5\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
$c_{pw}$ & Specific heat capacity of water vapor at constant pressure & $1930.0\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
$c_{vd}$ & Specific heat capacity of dry air at constant volume & $717.5\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
$c_{vw}$ & Specific heat capacity of water vapor at constant volume & $1460.0\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
$R_d$ & Gas constant for dry air & $287.0\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
$R_w$ & Gas constant for water vapor & $461.5$ J kg$^{-1}$ K$^{-1}$ \\
%$\kappa$ & Ratio of $R_d$ to $c_p$ & $2/7$ \\
$\varepsilon$ & Ratio of $R_d$ to $R_w$ & $0.622$ \\
$M_v$ & Constant for virtual temperature conversion & $0.608$ \\
$\rho_{water}$ & Reference density of water & 1000 kg m$^{-3}$ \\
\hline 
\end{tabular*}

@@

\section{Moist Nonhydrostatic Equation Sets} \label{sec:EquationSets}

In this section we describe the fluid equations utilized by nonhydrostatic models.  The material derivative is used for  quantities in the Lagrangian frame (following individual air parcels), and is given by
\begin{align}
\diff{}{t} = \pdiff{}{t} + \vb{u} \cdot \nabla.
\end{align}  Note that tracer variables $q_i$, including specific humidity $q$, satisfy the simple Lagrangian relationship
\begin{align} \label{eq:TracerTransport}
\diff{q_i}{t} = 0.
\end{align}

\subsection{Diagnostic relationships}

The atmospheric fluid is assumed to be an ideal gas.  For moist air, the ideal gas constant $R^\ast$, specific heat capacity at constant pressure $c_p^\ast$ and specific heat capacity at constant volume $c_v^\ast$ are given by
\begin{align}
R^\ast =& R_d + (R_w - R_d) q, & c_p^\ast =& c_{pd} + (c_{pw} - c_{pd}) q, & c_v^\ast =& c_{vd} + (c_{vw} - c_{vd}) q.
\end{align}  Note that in many models, $R^\ast$, $c_p^\ast$ and $c_v^\ast$ are approximated by $R_d$, $c_{pd}$ and $c_{vd}$, respectively.  Dry air, water vapor and moist air quantities all satisfy the linear relationship $R = c_p - c_v$.  For a two-fluid system (dry air plus water vapor), two independent variables plus the specific humidity $q$ are needed to describe the thermodynamic state of the system.  Key thermodynamic variables include dry air density $\rho_d$, moist density $\rho$, pressure $p$, vapor pressure $e$, temperature $T$, virtual temperature $T_v$, Exner pressure $\pi$, potential temperature $\theta$, and virtual potential temperature $\theta_v$.  Common ratios $\kappa = R^\ast / c_p^\ast$, $\epsilon = R_d / R_w$, and $\gamma = c_p^\ast / c_v^\ast$ are adopted here.

Relationships between key thermodynamic variables arise from the ideal gas law, along with definitions of Exner pressure, potential temperature and virtual potential temperature
\begin{align} \label{eq:DiagnosticRelationships1}
p =& \rho R_d T_v, & \pi =& \left( \frac{p}{p_0} \right)^{\kappa}, & \theta =& T \left( \frac{p_0}{p} \right)^{\kappa}, & \theta_v =& T_v \left( \frac{p_0}{p} \right)^{\kappa},
\end{align}  which further give rise to
\begin{align}
p =& \left( \frac{\rho R_d \theta_v}{p_0^\kappa} \right)^{\kappa - 1}, & \pi =& \left( \frac{\rho R_d \theta_v}{p_0} \right)^{- R^\ast / c_v^\ast}, & \theta =& \frac{T}{\pi}, & \theta_v =& \frac{T_v}{\pi}.
\end{align}

Note that virtual temperature is typically approximated by
\begin{align}
T_v \approx& T \left(1 + \frac{(1-\epsilon)}{\epsilon} q \right),
\end{align} which arises from the exact relationship
\begin{align}
T_v = \frac{T}{1 - \frac{e}{p} (1 - \epsilon)},
\end{align} upon applying $e/p = q / \epsilon$ and using a Taylor expansion around $q = 0$.

\subsection{Prognostic equations for thermodynamic variables}

Note that, as a consequence of (\ref{eq:TracerTransport}), the following simplifications can be applied:
\begin{align}
\frac{1}{T_v} \diff{T_v}{t} =& \frac{1}{T} \diff{T}{t}, & \diff{R^\ast}{t} =& 0, & \diff{c_p^\ast}{t} =& 0, & \diff{c_v^\ast}{t} =& 0.
\end{align}

Mass conservation is typically represented through the continuity equation, which can be written in the Lagrangian frame as
\begin{align} \label{eq:ContinuityEquation}
\diff{\rho}{t} =& - \rho \nabla \cdot \vb{u},
\end{align} or equivalently in the Eulerian frame,
\begin{align}
\pdiff{\rho}{t} =& - \nabla \cdot (\rho \vb{u}).
\end{align}

Further prognostic relationships can be derived from the thermodynamic equation, including the diabatic heating rate $J$,
\begin{align}
\frac{1}{T} \diff{T}{t} - \frac{\kappa}{p} \diff{p}{t} =& \frac{J}{c_p^\ast},
\end{align} which can be alternatively written as
\begin{align}
\diff{\theta}{t} = \frac{J \theta}{c_p^\ast}, \quad \mbox{or} \quad \diff{\theta_v}{t} = \frac{J \theta_v}{c_p^\ast}.
\end{align}  These equations can then be combined with (\ref{eq:ContinuityEquation}) to obtain
\begin{align}
\pdiff{}{t} (\rho \theta_v) + \nabla \cdot (\rho \theta_v \vb{u}) = \frac{J \rho \theta_v}{c_p^\ast},
\end{align} or similarly for $\theta$.  In conjunction with the material derivative of the ideal gas law,  
\begin{align}
\frac{1}{p} \diff{p}{t} = \frac{1}{\rho} \diff{\rho}{t} + \frac{1}{T_v} \diff{T_v}{t},
\end{align} the thermodynamic equation can be written in the form
\begin{align}
\frac{c_v^\ast}{R^\ast T_v} \diff{T_v}{t} - \frac{1}{\rho} \diff{\rho}{t} = \frac{J}{R^\ast}.
\end{align}  Then substituting (\ref{eq:ContinuityEquation}) gives a prognostic equation for virtual temperature,
\begin{align}
\frac{c_v^\ast}{R^\ast} \diff{T_v}{t} + T_v \nabla \cdot \vb{u} = \frac{J T_v}{R^\ast}.
\end{align}  The prognostic equation for temperature is identical except with $T$ substituted for $T_v$.  An analogous equations for pressure can be obtained through a similar procedure,
\begin{align} \label{eq:PressurePrognosticEq}
\frac{c_v^\ast}{c_p^\ast} \diff{p}{t} + p \nabla \cdot \vb{u} = \frac{J p}{c_p^\ast}.
\end{align}  And similarly for Exner pressure,
\begin{align}
\frac{c_v^\ast}{R^\ast} \diff{\pi}{t} + \pi \nabla \cdot \vb{u} = \frac{J \pi}{c_p^\ast}.
\end{align}

\subsection{Momentum Equations}

In coordinate-invariant form the prognostic velocity equations may be written in either the Lagrangian or Eulerian frame as
\begin{align} \label{eq:PrognosticVelocity1}
\diff{\vb{u}}{t} = \pdiff{\vb{u}}{t} + \vb{u} \cdot \nabla \vb{u} = - \frac{1}{\rho} \nabla p - 2 \vg{\Omega} \times \vb{u} - \nabla \Phi,
\end{align} where $\vg{\Omega}$ denotes the planetary vorticity vector and $\Phi$ is the geopotential function.  The three terms on the right-hand-side of this expression correspond to pressure gradient, Coriolis, and gravitational force, respectively.  In Eulerian form one must be careful with the treatment of the momentum advection term $\vb{u} \cdot \nabla \vb{u}$, since for an arbitrary coordinate frame will give rise to Christoffel symbols associated with derivatives of the vector basis {\color{red}[citation needed]}.  Note that it is common to rewrite the pressure gradient force using the relationship
\begin{align}
- \frac{1}{\rho} \nabla p = - c_p^\ast \theta_v \nabla \pi,
\end{align} which follows from (\ref{eq:DiagnosticRelationships1}).  A second form of (\ref{eq:PrognosticVelocity1}) emerges on substituting the vector calculus identity
\begin{align}
\vb{u} \cdot \nabla \vb{u} =& \nabla K + \vg{\zeta} \times \vb{u},
\end{align} where $K = \frac{1}{2} (\vb{u} \cdot \vb{u})$ is the 3D Kinetic energy and $\vg{\zeta} = \nabla \times \vb{u}$ is the 3D relative vorticity vector.  This gives rise to the vector-invariant form,
\begin{align} \label{eq:PrognosticVelocity2}
\pdiff{\vb{u}}{t} = - \frac{1}{\rho} \nabla p - (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} - \nabla (K + \Phi). 
\end{align}  Because no gradients of vectors appear in this equation, it avoids derivatives of the coordinate basis that would emerge in (\ref{eq:PrognosticVelocity1}).  In conjunction with (\ref{eq:ContinuityEquation}), both (\ref{eq:PrognosticVelocity1}) also give rise to the flux-form momentum equations,
\begin{align}
\pdiff{}{t} (\rho \vb{u}) = - \nabla \cdot (\vb{u} \otimes \vb{u} + \mathcal{I} p) - 2 \vg{\Omega} \times (\rho \vb{u}) - \rho \nabla \Phi,
\end{align} where $\vb{u} \otimes \vb{u}$ denotes the outer product and $\mathcal{I}$ is the identity matrix.

The equations above still provide some flexibility with regards to the choice of $\Phi$ and $\vg{\Omega}$.  For \textit{deep atmosphere} models, one typically chooses
\begin{align}
\Phi = \frac{g r^3}{3 a^2}, \quad \mbox{and} \quad \vg{\Omega} = \Omega ( \vb{k} \sin \phi + \vb{j} \cos \phi),
\end{align} where $g$ is gravitational acceleration at the surface, $r$ is the radial coordinate, $a$ is the radius of the planet, $\Omega$ is the rotation rate (in s$^{-1}$), $\phi$ is the latitude, $\vb{j}$ is the unit vector oriented in the meridional direction, and $\vb{k}$ is the unit vector oriented in the vertical direction.  For \textit{shallow atmosphere} models, these expressions take the simpler form
\begin{align}
\Phi = g z, \quad \mbox{and} \quad \vg{\Omega} = \Omega \sin \phi \vb{k},
\end{align} where $z$ is the altitude above the surface.  In this case we write $2 \vg{\Omega} = f \vb{k}$, where $f = 2 \Omega \sin \phi$ is the Coriolis parameter.

\subsection{Conservation of Energy}

\subsection{Discretization}

Horizontal momentum equations:

Lagrangian form:
\begin{align}
\diff{\vb{u}_h}{t} + f \vb{k} \times \vb{u}_h + \frac{1}{\rho} \nabla_s p + \nabla_s \phi = 0.
\end{align}

Clark form:
\begin{align}
\pdiff{\vb{u}_h}{t} + \nabla_h (K + \phi) + (\zeta + f) \vb{k} \times \vb{u}_h + \ldots + \dot{s} \nabla_h \left( \pdiff{z}{s} w \right)
\end{align}

Vertical velocity:
\begin{align}
\diff{w}{t} = 
\end{align}

\subsection{Discretization}

The equations in this section make no assumptions about the horizontal coordinate frame, and are written in terms of an arbitrary vertical coordinate $s(t,x,y,z)$.  In this context, the material derivative takes the form
\begin{align}
\diff{}{t} = \pdiff{}{t} + \vb{u}_h \cdot \nabla_s + \dot{s} \pdiff{}{s},
\end{align} where $\vb{u}_h$ is the 2D horizontal velocity vector, $\nabla_s$ is the horizontal gradient operator along surfaces of constant $s$, and $\dot{s}$ denotes the flow velocity perpendicular to surfaces of constant $s$, which relates the physical velocity components $\vb{u}_h$ and $w$ via
\begin{align}
\dot{s} = \diff{s}{t} = \pdiff{s}{t} + \vb{u}_h \cdot \nabla_z s + w \pdiff{s}{z},
\end{align} where $\nabla_z$ is the horizontal gradient operator along surfaces of constant $z$.

@@

\noindent with horizontal velocity ${\bf V}_h$, 
generalized vertical velocity $\dot{ \zeta }=\frac{d\zeta}{dt}$, 
vertical velocity $w$,
geopotential $\phi$, 
virtual temperature $T_v$, 
pressure $p$ and
hydrostatic pressure $\pi$. The variable 
$\mu$ is a measure of departure from hydrostatic balance, 
$s$ is related to the surface pressure $\pi_s$ and $B$ is a metric term. The last two variables are defined in (\ref{eq:gem_zeta}).

@@

@@

\subsection{Tracer Transport}
